An Intrinsically Typed Solution for the List Machine Benchmark

### 2020-06-21 papers/SBLP2020/paper.tex
Finished a first version of comparison section.

\title{A Certified Interpreter for the List Machine Benchmark}

@@

We implemented 14 tasks from the list machine benchmark in the Agda programming language.
The total number of lines for each task is summarized in the next table.

\begin{table}[!htb]
\begin{tabular}{rl|rrr}
    & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
1.  & Operational Semantics                        & 126                       & 98                      & 97                       \\
2.  & Derive $p \Downarrow$                        & 1                         & 8                       &                          \\ \hline
3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
4.  & $\sqcap$ algorithm                           & *                         & *                       & 13                       \\
5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & no                      & 1                        \\
6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
7.  & Prove properties of $\sqcap$                 & 114                       & 21                      & 124                      \\
8.  & State soundness theorem                      & 29                        & 15                      & *                        \\
9.  & Prove soundness of $\vDash_{\textrm{prog}} p : \Pi$
                                                   & 2060                      & 315                     & *                        \\ \hline
10. & Efficient algorithm                          & 22                        & 145                     & 98                       \\
11. & Derive $\vdash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & 1                       & 1                        \\
12. & Prove termination of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 18                        & 0                       & 0                        \\
13. & Scalable type-checker                        & yes                       & yes                     & yes                      \\
14. & Prove soundness of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 347                       & 141                     & * 
\end{tabular}
\end{table}

The total time for parsing and proof checking our Agda implementation was around 10 seconds 
on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5.

@@

>>>>

\title{An Intrinsically Typed Solution for the List Machine Benchmark}

@@

We implemented 14 tasks from the list machine benchmark in the Agda programming language.
The selected tasks considered by us are the ones done by Appel et. al.~\cite{Appel07}.
The next table summarize the total number of lines for our results together with Appel's.

\begin{table}[!htb]
\begin{tabular}{rl|rrr}
    & Task                                         & \multicolumn{1}{l}{Twelf} & \multicolumn{1}{l}{Coq} & \multicolumn{1}{l}{Agda} \\ \hline
1.  & Operational Semantics                        & 126                       & 98                      & 109                      \\
2.  & Derive $p \Downarrow$                        & 1                         & 8                       & -                        \\ \hline
3.  & Type System $\vDash_{\textrm{prog}} p : \Pi$ & 167                       & 130                     & 62                       \\
4.  & $\sqcap$ algorithm                           & *                         & *                       & 13                       \\
5.  & Derive $\vDash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & no                      & -                        \\
6.  & State properties of $\sqcap$                 & 12                        & 13                      & 6                        \\
7.  & Prove properties of $\sqcap$                 & 114                       & 21                      & 124                      \\
8.  & State soundness theorem                      & 29                        & 15                      & *                        \\
9.  & Prove soundness of $\vDash_{\textrm{prog}} p : \Pi$
                                                   & 2060                      & 315                     & *                        \\ \hline
10. & Efficient algorithm                          & 22                        & 145                     & 98                       \\
11. & Derive $\vdash_{\textrm{prog}} p_{\textrm{sample}} : \Pi_{\textrm{sample}}$
                                                   & 1                         & 1                       & 1                        \\
12. & Prove termination of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 18                        & 0                       & 0                        \\
13. & Scalable type-checker                        & yes                       & yes                     & yes                      \\
14. & Prove soundness of $\vdash_{\textrm{prog}} p : \Pi$
                                                   & 347                       & 141                     & * 
\end{tabular}
\end{table}

The total time for parsing and proof checking our Agda implementation was around 10 seconds 
on a machine with a Intel Core I7 1.7 GHz, 8GB RAM running Mac OS X 10.15.5. We briefly comment 
on our Agda coding of these 14 tasks.

\begin{enumerate}
  \item \textbf{Operational semantics.} Instead of using a inductive type for representing the operational semantics, we choose to
        use a definitional interpreter for our intrinsincally-typed representation. Our implementation for the operational
        semantics is composed by 38 lines for the type syntax and 71 for the definitional interpreter.
  \item \textbf{Derive $p\,\Downarrow$.} Since we used an definitional intepreter for representing the semantics, we can derive
        $p\,\Downarrow$ just by executing the interpreter on $p$.
  \item \textbf{Represent the type system.} Our type system representation consists of the intrinsically-typed encoding of
        the list machine programs and it is already counted in as part of the operational semantics of our solution.
  \item \textbf{Least common supertype algorithm.} In order to implement this task, we specified the least common super type
        as a relation and implemented the algorithm which shows that this relation is indeed a function. In this way, we
        guarantee its soundness w.r.t. its specification.
  \item \textbf{Derive an example of type checking.} Our approach to build a derivation for a sample list machine program
        is just execute the type checker over it. Since our type checker returns an intrinsincally-typed representation of
        the input and such typed syntax is equivalent to the type system, it can be used to build type derivations.
  \item \textbf{State properties of the least common supertype.} This is trivial, since it is just a matter to encode the
        desired properties as Agda function types.
  \item \textbf{Prove properties of the least common supertype.} Proofs of all proved properties about the least common supertype
        are simple recursive functions over the least common supertype and subtyping relations definition.
  \item \textbf{State the soundness theorem for the type system.} Following Amin et. al.~\cite{Amin17}, we represent the soundness
        theorem statement as the type of our definitional interpreter for the intrinsically-typed syntax for the list machine
        programs.
  \item \textbf{Prove the soundness theorem for the type system.} Following Amin et. al.~\cite{Amin17}, the proof of the soundness
        theorem is just the implementation of the definitional interpreter. Soundness is ensured by construction since the interpreter
        produces, as result, an environment of well-typed values resulting from the execution of the input program.
  \item \textbf{Asymptotically efficient algorithm.} Our current implementation of the typechecker takes quadratic time. The reason
        for this inneficiency is the representation of environments as lists / vectors. The use of better data structures (like finite
        maps implemented by efficient search trees) can solve this issue. We leave this fix to future work.
  \item \textbf{Simulate the new algorithm.} This task is just one line of code, since it ammounts to calling the typechecker
        on the input program.
  \item \textbf{Prove the termination of the type checker.} This task is trivial in our setting, since all defined Agda functions
        must be total. The totality is ensured by Agda's termination / totality checker.
  \item \textbf{Scalable type-checker.} Agda code can be compiled to machine code using its GHC-Haskell backend. Since GHC is
        an industrial strength compiler, the backend can generate an efficient executable for the machine interpreter and typechecker. 
  \item \textbf{Prove soundness of type-checker.} In our approach, the soundness of the typechecker is ensured by construction,
        since it returns the intrinsically-typed representation of the input program which corresponds to its typing derivation.
\end{enumerate}

@@

\paragraph{Benchmarks for PL mechanization}{
POPLMark chalenge~\cite{Aydemir05} .... Another chalenge was proposed by Pientka et. al.~\cite{Pientka18}...
}

\paragraph{Definitional intepreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
}
